The ChEMBL database as Linked Open Data

### 2012-07-09 article.tex
Workaround for a stupid bug in TexMaker that does not find the end of the section title properly
Tuned the intro and methods section

\section*{Introduction}\label{s1}

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

ChEMBL contains descriptions for biological activities involving over a million chemical 
entities, extracted from literature. This provides a unique resource for drug researchers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. 
The ChEMBL dataset is available for download and can be browsed using a web interface. 
The former requires scientists to import the data into a relational database, while the 
latter limits the machine access to the data. Two independent teams have previously mapped 
the ChEMBL dataset to RDF, Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}.

This paper presents the current state of the ChEMBL-RDF dataset, details of the latest structures
and ontologies used to map version 13 of ChEMBL to RDF, along with new example use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

\section*{Methods}\label{s2}

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL datadump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest Group~\cite{Marshall2012}.

In the previous RDF conversion the RDF triples in ChEMBL-RDF were represented using properties and c
lasses from a custom, ad-hoc ontology~\cite{Willighagen2011}. The current version used community-proposed
ontologies, making the RDF more interoperable. The ChEMBL-RDF 13 dataset uses ontology
standards such as the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology~\cite{Hastings2011} for literature
references, and domain ontologies like the Protein Ontology\cite{Sidhu2006} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper, various shorthand prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable using the Kasabi platform~\cite{kasabi}. The Kasabi hosting provides a SPARQL
end point which requires a user account. A user account can be free for casual usage, or paid for consistent usage.

@@

\section*{Results}\label{s3}

>>>>

\section*{Introduction}
\label{s1}

The current scientific data deluge, in which datasets grow faster in size and complexity than scientists can handle,
provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{Staab2007}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

ChEMBL contains descriptions for biological activities involving over a million chemical 
entities, extracted from literature. This provides a unique resource for drug researchers~\cite{Gaulton2012,Warr2009}.
It is updated on a fairly frequent basis as the existing data is further curated and new data is added. 
The ChEMBL dataset is available for download and can be browsed using a web interface. 
The former requires scientists to import the data into a relational database, while the 
latter limits the machine access to the data.

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped  the ChEMBL dataset to the Resource Description Framework (RDF),
Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small, so-called triples reflecting very basic facts. Things
in RDF are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not any, domain, making it a perfect
approach for data integration in the life sciences.

This paper presents an update on the ChEMBL-RDF dataset, and details of the latest structures
and ontologies used to map ChEMBL 13 to RDF, along with new use cases, showing 
links to other datasets using their RDF URIs to further support research in the life sciences.

\section*{Methods}
\label{s2}

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL datadump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest
Group~\cite{Marshall2012}. This process has not changed much.

The previous ChEMBL-RDF conversion used, however, custom predicates and classes to represent RDF triples, defining
an ad-hoc ontology~\cite{Willighagen2011}. The current version, instead, uses community-proposed
ontologies, making the RDF more interoperable. The new triple dataset uses standard ontologies
including the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology~\cite{Hastings2011} for literature
references and citations, and domain ontologies like the Protein Ontology\cite{Sidhu2006} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper, various prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable using the Kasabi platform~\cite{kasabi}. The Kasabi hosting provides a SPARQL
end point which requires a user account. A user account can be acquired for free for casual usage,
or paid for uncapped  use.

@@

\section*{Results}
\label{s3}
